Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2022-10-27 analysis/paper_draft3.Rmd
added figure of Batson workflow; discussion of testing for alpha sensitivity; right to attend criminal trials

Since _Batson v. Kentucky_ [-@batson1986], a party violates the Equal Protection Clause of the US Constitution by using peremptory challenges if motivated by the prospective jurors' race, ethnicity, sex, or certain other cognizable classes. The party bringing a _Batson_ challenge bears the burden of proving such illegal bias is more likely than not to be true [@LaFave2022, $\S$ 22.3(d)]. For similar challenges under State law, a few States require only that an "objective observer" or an "objectively reasonable person" would find that race, ethnicity or another cognizable-class was a "factor" in that party's use of strikes [Wash. General Rule. 37(e); Calif. Code of Civil Procedure $\S$ 231.7(d)(1)]. Peremptory-challenge procedure, and thus the task of proving such illegal bias, varies not only by State, but also by court, including the number of peremptory challenges assigned to each side and the order in which each party uses those strikes [@nscs-jury2018; @Williams2017].

In any case, evidence of such illegal bias may include data on a party's use of peremptory challenges in past cases [@flowers2019, 2243; Wash. General Rule 37(g)(v); Calif. Code Civil Procedure $\S$ 231.7(d)(3)(G)]. Prior studies have collected historical strike data in past cases and reported the observed difference in strike rates by race or sex. Typically, they test for the probability of observing a non-zero difference in strike rates by the race or sex of the struck prospective jurors, given repeated sampling from a hypothetical population of peremptory strikes with zero such difference [e.g., @Eisenberg2017; @Grosso2012; for discussion, see @Gastwirth2014, 289-297; @Gastwirth2013]. Prior studies have also modeled how much a prospective juror's race affected the odds of being struck. For different modeling approaches using the same historical strike data from Mississippi, see @Craft2018a; @DeCamp2021; and @Dunn2021.

@@

Following others (e.g., @Kadane2018a; @Barrett2007; @Gastwirth2005, 183], we model a peremptory-challenge process in which each party strikes prospective jurors in an alternating sequence. Under such a procedure, the trial judge rules on all challenges for cause before the parties exercise any peremptories. Then, of the potential jurors who remain, a subset of them are subject to peremptory strike, usually a number that corresponds to the number seats on the jury (plus alternates, if any). The parties exercise their strikes on anyone among this subset of potential jurors in an alternating sequence. Once all strikes are used or waived, the remaining prospective jurors are each assigned to seats on the jury plus a few alternate jurors.

@@

Moreover, with a tool like our prototype app, attorneys can use the approach here in to detect strike bias for _Batson_ and related challenges, in three ways. First, the posterior distribution of the bias parameter can help the lawyer decide in real time whether it is worth raising a  _Batson_ challenge at all.

```{r batsonworkflowfig}

library(DiagrammeR)
grViz(diagram = "digraph flowchart {
  rankdir = LR
  node [fontname = arial, shape = box]
  tab1 [label = '@@1']
  tab2 [label = '@@2']
  tab3 [label = '@@3']

Second, if raised, the tool can help the _Batson_ challenger meet its evidentiary burdens at key steps of a _Batson_ challenge. A _Batson_ challenge typically proceeds in a three-step sequence (Figure \@ref(fig:batsonworkflowfig)). First, the challenger must meet an initial burden (the _prima facie_ case) of producing just enough evidence of strike bias to raise a sufficient inference of illegal bias. If so, then at step two, the striking attorney must produce (permissible) reasons for their challenged strikes. Third, the challenger must discredit those proffered reasons and ultimately persuade the trial judge to find that it is more likely than not that the striking attorney acted with illegal bias.

To meets its evidentiary burdens at the first and third steps, the attorney can introduce the striking attorney's estimated bias parameter as relevant statistical evidence. In this way, the tool helps detect bias. But like any statistical evidence submitted for a _Batson_ challenge, it cannot alone be the basis of deciding a _Batson_ challenge. That is because it alone cannot account for the veracity of the striking attorney's reasons offered at step two and because the _Batson_ procedure allows for any type of circumstantial evidence of bias at steps one and three.

@@

Second, even with complete past strike data (no missingness), incompatibility may occur if attorney bias depends on certain trial-level characteristics (e.g., defendant race, charge severity) that take on one value in the current trial but that vary across the historical trials. For example, suppose an attorney who is more likely to strike Black prospective jurors because the defendant is Black; who has zero strike bias in cases with white defendants; and whose historical strike data comes from past trials with an equal number of Black and white defendants. If so, then bias detection in the current trial with a Black defendant may depend in part on whether we use all the historical data or only the subset of past trials with Black defendants.

@@

>>>>

Since _Batson v. Kentucky_ [-@batson1986], a party violates the Equal Protection Clause of the US Constitution by using peremptory challenges if motivated by the prospective jurors' race, ethnicity, sex, or certain other cognizable classes. The party bringing a _Batson_ challenge bears the burden of proving such illegal bias is more likely than not to be true [@LaFave2022, $\S$ 22.3(d)].

```{r batsonworkflowfig, fig.cap="Batson workflow"}

library(DiagrammeR)
grViz(diagram = "digraph flowchart {
  rankdir = LR
  node [fontname = arial, shape = rectangle]
  tab1 [label = '@@1']
  tab2 [label = '@@2']
  tab3 [label = '@@3']

A _Batson_ challenge typically proceeds in a three-step sequence (Figure \@ref(fig:batsonworkflowfig)). First, the challenger must meet an initial burden (the _prima facie_ case) of producing just enough evidence of strike bias to raise a sufficient initial inference of illegal bias. If so, then at step two, the striking attorney must produce (permissible) reasons for their challenged strikes. Third, the challenger must discredit those proffered reasons and ultimately persuade the trial judge to find that it is more likely than not that the striking attorney acted with illegal bias.

For similar challenges under State law, a few States require only that an "objective observer" or an "objectively reasonable person" would find that race, ethnicity or another cognizable-class was a "factor" in that party's use of strikes [Wash. General Rule 37(e); Calif. Code of Civil Procedure $\S$ 231.7(d)(1); Conn. Superior Court Rule 5-12(d)]. Peremptory-challenge procedure, and thus the task of proving such illegal bias, varies not only by State, but also by court, including the number of peremptory challenges assigned to each side and the order in which each party uses those strikes [@nscs-jury2018; @Williams2017].

In any case, evidence of such illegal bias may include data on a party's use of peremptory challenges in past cases [@flowers2019, 2243; Wash. General Rule 37(g)(v); Calif. Code Civil Procedure $\S$ 231.7(d)(3)(G)]. Prior studies have collected historical strike data in past cases and reported the observed difference in strike rates by race or sex. Typically, they test for the probability of observing a non-zero difference in strike rates by the race or sex of the struck prospective jurors, given repeated sampling from a hypothetical population of peremptory strikes with zero such difference [e.g., @Eisenberg2017; @Grosso2012; for discussion, see @Gastwirth2014, 289-297; @Gastwirth2013]. Prior studies have also modeled how much a prospective juror's race affected the odds of being struck. For different modeling approaches using the same historical strike data from Mississippi, see @Craft2018a; @DeCamp2021; and @Dunn2022.

@@

Following others [e.g., @Kadane2018a; @Barrett2007; @Gastwirth2005, 183], we model a peremptory-challenge process in which each party strikes prospective jurors in an alternating sequence. Under such a procedure, the trial judge rules on all challenges for cause before the parties exercise any peremptories. Then, of the potential jurors who remain, a subset of them are subject to peremptory strike, usually a number that corresponds to the number seats on the jury (plus alternates, if any). The parties exercise their strikes on anyone among this subset of potential jurors in an alternating sequence. Once all strikes are used or waived, the remaining prospective jurors are each assigned to seats on the jury plus a few alternate jurors.

@@

Moreover, with a tool like our prototype app, attorneys can use the approach here in to better detect strike bias for _Batson_ and related challenges, in three ways. First, the posterior distribution of the bias parameter can help the lawyer decide in real time whether it is worth raising a  _Batson_ challenge at all.

Second, if raised, the tool can help the _Batson_ challenger meet its evidentiary burdens at steps one and three of a _Batson_ challenge (Figure \@ref(fig:batsonworkflowfig)). To meets its evidentiary burdens at the first and third steps, the attorney can introduce the posterior distribution of the striking attorney's bias parameter as relevant statistical evidence. In this way, the tool helps detect bias. But it in no way suffices for deciding a _Batson_ challenge. Such statistical evidence alone cannot account for the veracity of the striking attorney's reasons offered at step two. Besides, _Batson_ procedure allows for any type of circumstantial evidence of bias at steps one and three.

@@

At the same time, US law currently affords an avenue for reducing this missing data problem by systematically collecting historical strike data prospectively. A criminal defendant has a Sixth Amendment right to a public trial that presumptively precludes excluding the public from the voir dire of prospective jurors. Similarly, the public has a presumptive First Amendment right to attend jury voir dire in criminal cases, even if prosecutor and defendant agree to exclude the public [@press1984; @presley2010]. Upon exercise of these rights, courtroom observors can attend jury voir dire and collect relevant strike data based on what they observe in real time.  Even if historical strike data is hard to collect retrospectively, courtroom observers can over time build up a dataset for strike data that they would have been less likely to obtain from court records after the fact.

Second, even with complete strike data (no missingness), incompatibility may occur if attorney bias depends on certain trial-level characteristics (e.g., defendant race, charge severity) that take on one value in the current trial but that vary across the historical trials. For example, suppose an attorney who is more likely to strike Black prospective jurors because the defendant is Black; who has zero strike bias in cases with white defendants; and whose historical strike data comes from past trials with an equal number of Black and white defendants. If so, then bias detection in the current trial with a Black defendant may depend in part on whether we use all the historical data or only the subset of past trials with Black defendants.

@@

More importantly, in cases where lawyers contest the degree of compatibility, those lawyers should interrogate the choice of $\alpha$ by selecting multiple $\alpha$ values and observing how sensitive the posterior distribution of the striking attorney's bias parameter is to the choice of $\alpha$ value. (Our protoype app permits this kind of testing of sensitivity in real time by providing multiple options for $\alpha$ values.) If the inference of bias is sensitive to the choice of $\alpha$ values, all else equal, then worries about compatibility are warranted and the historical strike data, without more, is less helpful in steps one and three of the _Batson_ procedure than it would otherwise would be. One virtue of this approach is that the choice of $\alpha$ explicitly denotes the degree of weight placed on the historical strike data in the analysis.
